The impact of early massive mergers on the chemical evolution of Milky Way-like galaxies: insights from NIHAO-UHD simulations

### 2023-04-04 src/tex/ms.tex
[showyourwork] overleaf sync

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, one prerequisite to follow the internal disk dynamics and suppress spurious heating as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) is enough resolution ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019,Ludlow2021}). At the same time a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is needed to accurately capture the star's birth kinematics (e.g. \citealt{Ratcliffe2022, Lu2022}) and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013, Buck2019b}).

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2018, Font2020, Applebaum2021, Agertz2021, Bird2021, Khoperskov2022-InSitu}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's evolution \citep[e.g.][]{Bignone2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022a, Rey2022}.

@@

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2022b} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST DR7 combined with {\it Gaia} eDR3. When evaluating the chemical evolution across birth radii and lookback time using APOGEE DR17 red giant branch disk stars, Ratcliffe et al. submitted reported a similar steepening in the metallicity gradient at $\sim9$ Gyr ago, with a similar trend also appearing in other [X/H] abundances. Here we use the NIHAO-UHD suite of cosmological hydrodynamical simulations of MW-mass galaxies \citep{Buck2020a} %in order 
to investigate the influence of early merger events on the chemo-dynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. We focus especially on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is structured as follows: In \S~2 we describe the simulations on which we base our analysis, followed by a presentation of the impact of early massive mergers on the metallicity gradient of the cold gas disk in \S~3. In \S~4 we turn to analyse where merger stars and gas end up in the age-metallicity relation  and the [$\alpha$/Fe] vs. [Fe/H] plane of the main galaxy. In \S~5 and \S~6 we summarize and conclude our findings.

@@

For this work we use four simulations from the NIHAO-UHD suite \citep{Buck2020a} part of the Numerical Investigation of a Hundred Astronomical Objects (NIHAO) simulation suite \citep{Wang2015}. Parts of this simulation suite have previously been used to study the build-up of MW's peanut-shaped bulge \citep{Buck2018,Buck2019}, investigate the stellar bar properties \citep{Hilmi2020}, infer the MW's dark halo spin \citep{Obreja2022}, study the dwarf galaxy inventory of MW mass galaxies \citep{Buck2019a} or investigate the age-metallicity relation of MW disk stars \citep{Lu2022} including the chemical bimodality of disk stars \citep{Buck2020}, their abundances \citep{Lu2022a} and the origin of very metal-poor stars inside the stellar disk \citep{Sestito2021}.
Comparing the properties of these galaxies with observations of the MW and local disk galaxies from the SPARC data \citep{Lelli2016}, \citet{Buck2020a} showed that simulated galaxy properties agree well with observations.

>>>>

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, one prerequisite to follow the internal disk dynamics and suppress spurious heating as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) is enough resolution ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019,Ludlow2021}). At the same time a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is needed to accurately capture the star's birth kinematics (e.g. \citealt{Ratcliffe2022, Lu2022}) and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013, Buck2020a}).

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2020a, Font2020, Applebaum2021, Agertz2021, Bird2021, Khoperskov2022-InSitu}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's evolution \citep[e.g.][]{Bignone2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022a, Rey2022}.

@@

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2022b} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST DR7 combined with {\it Gaia} eDR3. When evaluating the chemical evolution across birth radii and lookback time using APOGEE DR17 red giant branch disk stars, Ratcliffe et al. submitted reported a similar steepening in the metallicity gradient at $\sim9$ Gyr ago, with a similar trend also appearing in other [X/H] abundances. Here we use the NIHAO-UHD suite of cosmological hydrodynamical simulations of MW-mass galaxies \citep{Buck2020a} to investigate the influence of early merger events on the chemo-dynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. We focus especially on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is structured as follows: In \S~2 we describe the simulations on which we base our analysis, followed by a presentation of the impact of early massive mergers on the metallicity gradient of the cold gas disk in \S~3. In \S~4 we turn to analyse where merger stars and gas end up in the age-metallicity relation  and the [$\alpha$/Fe] vs. [Fe/H] plane of the main galaxy. In \S~5 and \S~6 we summarize and conclude our findings.

@@

For this work we use four simulations from the NIHAO-UHD suite \citep{Buck2020a} part of the Numerical Investigation of a Hundred Astronomical Objects (NIHAO) simulation suite \citep{Wang2015}. Parts of this simulation suite have previously been used to study the build-up of MW's peanut-shaped bulge \citep{Buck2018,Buck2019b}, investigate the stellar bar properties \citep{Hilmi2020}, infer the MW's dark halo spin \citep{Obreja2022}, study the dwarf galaxy inventory of MW mass galaxies \citep{Buck2019} or investigate the age-metallicity relation of MW disk stars \citep{Lu2022} including the chemical bimodality of disk stars \citep{Buck2020}, their abundances \citep{Lu2022a} and the origin of very metal-poor stars inside the stellar disk \citep{Sestito2021}.
Comparing the properties of these galaxies with observations of the MW and local disk galaxies from the SPARC data \citep{Lelli2016}, \citet{Buck2020a} showed that simulated galaxy properties agree well with observations.
